---
title: "W271 Group Lab 2"
subtitle: "The Keeling Curve"
author: "Adam Kreitzman, Hailee Schuele, Lee Perkins, Paul Cooper"
output: bookdown::pdf_document2
fontsize: 11pt
geometry: margin=1in
---

```{r load packages, echo = FALSE, message = FALSE}
library(tidyverse)
library(tsibble)
library(forecast)
library(tseries)
# library(latex2exp)
# library(patchwork)
# library(lmtest)
library(ggplot2)
# library(car)
library(dplyr)
# library(nnet)
# library(GGally)
# library(gridExtra)
# library(magrittr)
library(lubridate)
library(feasts)
# library(sandwich)
library(fable)
# library(httr)
# library(readr)
# library(jsonlite)

set.seed(123)
theme_set(theme_minimal())
knitr::opts_chunk$set(dpi=1000)
```

```{r load data}
df <- co2 %>%
  as_tsibble(index = index)
```


# Part A - 1997


## Introduction

### Notes
Introduce the question to your audience. Suppose that they _could_ be interested in the question, but they don't have a deep background in the area. What is the question that you are addressing, why is it worth addressing, and what are you going to find at the completion of your analysis. Here are a few resource that you might use to start this motivation. 

- [Wikipedia](https://en.wikipedia.org/wiki/Keeling_Curve)
- [First Publication](./background/keeling_tellus_1960.pdf)
- [Autobiography of Keeling](./background/keeling_annual_review.pdf)


## CO2 Data

### Notes
- EDA
- How, where and why (https://gml.noaa.gov/ccgg/about/co2_measurements.html) the data is generated, 
- Investigation of the trend, seasonal and irregular elements. Trends both in levels and growth rates should be discussed (consider expressing longer-run growth rates as annualized averages).

### Content
```{r time series plot}
df %>%
  autoplot(value)
```

```{r acf plot}
df %>%
  ACF(value) %>%
  autoplot()
```

```{r pacf plot}
df %>%
  PACF(value) %>%
  autoplot()
```

```{r lag plot}
df %>%
  gg_lag(value, geom = "point", color = 1) +
  labs(title = "Lag Plot")
```



```{r seasonal plot}
df %>%
  gg_season(value, period = "year") +
  labs(title = "Annual Seasonal Plot")
```


```{r unit root test}
# DEL need to to an adf test also?

PP.test(df$value)
```


```{r growth rate}
# DEL how do I do this? Trends both in levels and growth rates should be discussed (consider expressing longer-run growth rates as annualized averages).
```


### Discussion
The time series plot shows an upward trend with seasonal components that have standard variance. This is supported by the lack of decline in the ACF, which indicates a trend. The slow decline in ACF could indicate an ARIMA process. The time series and ACF both point to a non-stationary process. The PACF also drops off rapidly after the first lag, which is characteristic of non-stationarity coming from a trend and seasonality. 

The lag plots are incredibly linear, which is a sign of seasonality. This seasonality can be seen in the annual seasonal plot. In this one, there seems to be a consistent increase in CO2 during the spring/early summer and a decrease in the early fall.

#### DEL is this correct? Doesn't look stationary
The Phillips Perron Test returns a significant p-value, which suggests there is stationarity. 

#### DEL what does this mean?
The PACF oscillates with a few significant lags here and there.


## Linear time trend model

### Notes
- Fit a linear time trend model to the `co2` series, and examine the characteristics of the residuals
- Compare this to a quadratic time trend model
- Discuss whether a logarithmic transformation of the data would be appropriate
- Fit a polynomial time trend model that incorporates seasonal dummy variables, and use this model to generate forecasts to the year 2020. 

### Content
```{r linear model}
linear_model <- df %>%
  model(lm = TSLM(value ~ index))

linear_model %>%
  report()
```

```{r linear residuals}
# Time vs Residuals line plot
linear_model %>%
  augment %>%
  autoplot(.resid) +
  labs(title = "Time Plot of Residuals - Linear Model")
```

```{r quadratic model}
# Cast index as a numeric variable
df$num_index <- as.numeric(df$index) - min(as.numeric(df$index))

# Fit quadratic model
quad_model <- df %>%
  model(lm = TSLM(value ~  poly(num_index, degree = 2, raw = TRUE)))

quad_model %>%
  report()

# DEL confirmed get same coefficients as below
# quad_model <- lm(value ~ poly(num_index, degree = 2, raw = TRUE), data = df)
```

```{r quadratic residuals}
# Time vs Residuals line plot
quad_model %>%
  augment %>%
  autoplot(.resid) +
  labs(title = "Time Plot of Residuals - Quadratic Model")
```

```{r fit seasonal model}
# Seasonal dummy variables
df <- df %>%
  mutate(month = lubridate::month(index))

seasonal_dummies <- model.matrix(~ factor(month) - 1, data = df)

colnames(seasonal_dummies) <- month.abb

df <- dplyr::bind_cols(df, seasonal_dummies)

# Fit polynomial model
# DEL should degree = 2?
poly_model <- df %>%
  model(lm = TSLM(value ~  poly(num_index, degree = 2, raw = TRUE) + 
                    Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov + Dec))

poly_model %>%
  report()
```


```{r poly forecast}
# Create a new data frame for forecasting
# DEL is all of this necessary?
new_data <- data.frame(num_index = max(df$num_index) + 1:276)

new_data$month <- 1:12
new_data$month <- rep(new_data$month, length.out = nrow(new_data))

new_data$values <- NA

seasonal_dummies <- model.matrix(~ factor(month) - 1, data = new_data)
colnames(seasonal_dummies) <- month.abb
new_data <- cbind(new_data, seasonal_dummies)

index_sequence <- seq.Date(as.Date(paste(1998, "01", "01", sep = "-")), as.Date(paste(2020, "12", "01", sep = "-")), by = "month")
new_data$index <- yearmonth(index_sequence)

df_forecast <- df
df_forecast <- dplyr::bind_rows(df_forecast, new_data)

# Forecast
forecast_poly_model <- forecast(poly_model, new_data = df_forecast)

forecast_poly_model %>%
  autoplot(df_forecast)
```


### Discussion
The residuals of the linear model do not appear to be white noise. The residuals of the quadratic model are closer to white noise.

Log transformations are generally used to stabalize variance, which does not seem to be needed here.


## ARIMA times series model 

### Notes
- Following all appropriate steps, choose an ARIMA model to fit to the series. 
- Discuss the characteristics of your model and how you selected between alternative ARIMA specifications. 
- Use your model (or models) to generate forecasts to the year 2022. 

### Content

```{r fit arima model}
# DEL confirm should include 1 (i.e. drift)
# DEL why doesn't it work when try to set PDQ values?
# DEL Vinod - For the purposes of the lab, you should focus on fitting an ARIMA model not a SARIMA model.

arima_model_aic <- df %>%
  model(ARIMA(value ~ 1 + pdq(1:10,1:2,1:10) + PDQ(0,0,0), ic = "aic", stepwise = FALSE, greedy = FALSE))

arima_model_aic %>%
  report()

# arima_model_aic_test <- df %>%
#   model(ARIMA(value ~ 1 + pdq(0:10,0:2,0:10) + PDQ(0:10,0:2,0:10), ic = "aic", stepwise = FALSE, greedy = FALSE))


arima_model_bic <- df %>%
  model(ARIMA(value ~ 1 + pdq(1:10,1:2,1:10) + PDQ(0,0,0), ic="bic", stepwise=FALSE, greedy=FALSE))

arima_model_bic %>%
  report()
```

```{r arima forecast}
arima_model <- df %>%
  model(ARIMA(value ~ 1 + pdq(2,1,4) + PDQ(0,0,0)))

forecast_arima_model <- forecast(arima_model, h = 288)

forecast_arima_model %>%
  autoplot(df)
```


### Discussion
Both the AIC and BIC models selected ARIMA(2,1,4). Differencing makes sense because it looked like there was non-stationarity. 


## Forecast atmospheric CO2 growth 

### Notes
- Generate predictions for when atmospheric CO2 is expected to be at 420 ppm and 500 ppm levels for the first and final times (consider prediction intervals as well as point estimates in your answer). 
- Generate a prediction for atmospheric CO2 levels in the year 2100. How confident are you that these will be accurate predictions?

### Content

```{r predictions}
# Predict 
forecast_arima_model_2100 <- forecast(arima_model, h = 5000)
# forecast_arima_model_2100

# DEL only doing point estimates, need to also take into account prediction interval
# First and last occurrence of 420ppm
first_420 <- min(forecast_arima_model_2100$index[forecast_arima_model_2100$.mean >= 420])
# DEL fix this
last_420 <- max(forecast_arima_model_2100$index[forecast_arima_model_2100$.mean >= 420])

first_420
last_420

# First and last occurrence of 500ppm
first_500 <- min(forecast_arima_model_2100$index[forecast_arima_model_2100$.mean >= 500])
# DEL fix this
last_500 <- max(forecast_arima_model_2100$index[forecast_arima_model_2100$.mean >= 500])

first_500
last_500

# tail(forecast_arima_model_2100, 1)$value
# tail(forecast_arima_model_2100)$value
```


```{r year 2100}
co2_2100 <- forecast_arima_model_2100 %>%
  filter(yearmonth(index) >= yearmonth("2100 Jan") & yearmonth(index) <= yearmonth("2100 Dec")) %>%
  summarise(mean_2100 = .mean)

co2_2100
```



### Discussion
Based on the newer science around climate change, specifically atmospheric levels of CO2 increasing exponentially, I'm not very confident in the estimates of this ARIMA model.

It's hard to get a "last" value because the variances are so wide the farther you go out



# Part B - The Present


## Introduction

### Notes
- briefly pose the question that you are evaluating, and describe what (if anything) has changed in the data generating process between 1997 and the present. 

### Discussion


## Data Pipeline for Mona Loa CO2 data

### Notes
- Conduct the same EDA on this data using the same, or very similar tools and views as you provided in your 1997 report. 
- Describe how the Keeling Curve evolved from 1997 to the present, noting where the series seems to be following similar trends to the series that you "evaluated in 1997" and where the series seems to be following different trends.

### Content
```{r read weekly mona loa data}
# Read in data
co2_present <- read.csv("co2_weekly_mlo.csv", comment.char = "#")

# Cast as tsibble
# DELETE figure out what the -999.99 average values mean
co2_present <- co2_present %>%
  mutate(time_index = yearweek(lubridate::make_datetime(year, month, day))) %>%
  filter(average >= 0) %>%
  as_tsibble(index = time_index)
```

```{r present time series plot}
co2_present %>%
  autoplot(average)
```

```{r present acf plot}
co2_present %>%
  tsibble::fill_gaps() %>%
  ACF(average) %>%
  autoplot()
```

```{r present pacf plot}
co2_present %>%
  tsibble::fill_gaps() %>%
  PACF(average) %>%
  autoplot()
```

```{r present lag plot}
co2_present %>%
  gg_lag(average, geom = "point", color = 1) +
  labs(title = "Lag Plot")
```



```{r seasonal plot}
co2_present %>%
  tsibble::fill_gaps() %>%
  gg_season(average, period = "year") +
  labs(title = "Annual Seasonal Plot")
```




### Discussion
Similar to the 1997 data,the time series plot shows an upward trend with seasonal components. Trends and seasonality are also supported by the ACF, PACF, and lag plot. The seasonal plot again shows increase in CO2 during the spring/early summer and a decrease in the early fall.

All in all, this is very similar to the data we saw in 1997. Perhaps even more linear.



## Compare linear model forecasts against realized CO2

### Notes
- Descriptively compare realized atmospheric CO2 levels to those predicted by your forecast from a linear time model in 1997 (i.e. "Task 2a"). (You do not need to run any formal tests for this task.) 

### Content
```{r linear forecast compare}
# Linear model predictions
# (2023 - 1997) years * 12 months = 312
forecast_linear_model <- forecast(linear_model, h = 312)

# Aggregate weekly data to monthly
# DELETE confirm this is correct, don't think it's calculating right
co2_present_agg <- co2_present %>%
  tsibble::index_by(yearmonth = yearmonth(time_index)) %>%
  summarise(monthly_average = mean(average)) %>%
  filter(yearmonth > yearmonth("1997 Dec"))

# Merge forecast data onto mone loa data
co2_present_agg <- merge(co2_present_agg, forecast_linear_model[, c("index", ".mean")],
                          by.x = "yearmonth", by.y = "index")
# Calculate residuals 
co2_present_agg <- rename(co2_present_agg, linear_monthly_average = .mean) %>%
  mutate(linear_residual = monthly_average - linear_monthly_average)

# Plot residuals
co2_present_agg %>%
  ggplot(aes(x = yearmonth, y = linear_residual)) +
  geom_point()
```

### Discussion



## Compare ARIMA models forecasts against realized CO2  

### Notes
- Descriptively compare realized atmospheric CO2 levels to those predicted by your forecast from the ARIMA model that you fitted in 1997 (i.e. "Task 3a"). 
- Describe how the Keeling Curve evolved from 1997 to the present. 

### Content
```{r arima forecast compare}
# ARIMA model predictions
forecast_arima_model <- forecast(arima_model, h = 312)

# Merge forecast data onto mone loa data
co2_present_agg <- merge(co2_present_agg, forecast_arima_model[, c("index", ".mean")],
                          by.x = "yearmonth", by.y = "index")
# Calculate residuals 
co2_present_agg <- rename(co2_present_agg, arima_monthly_average = .mean) %>%
  mutate(arima_residual = monthly_average - arima_monthly_average)

# Plot residuals
co2_present_agg %>%
  ggplot(aes(x = yearmonth, y = arima_residual)) +
  geom_point()
```

### Discussion


## Evaluate the performance of 1997 linear and ARIMA models 

### Notes
- How close was arima model to the truth about when CO2 would cross 420 ppm?
- Use the weekly data to generate a month-average series from 1997 to the present, and compare the overall forecasting performance of your models from Parts 2a and 3b over the entire period. (You should conduct formal tests for this task.) 

### Content
```{r 420 threshold}
min(co2_present_agg$yearmonth[co2_present_agg$monthly_average >= 420])
```
```{r formal test}
# DELETE What formal test should we use?
```


### Discussion



## Train best models on present data

### Notes
- Seasonally adjust the weekly NOAA data, and split both seasonally-adjusted (SA) and non-seasonally-adjusted (NSA) series into training and test sets, using the last two years of observations as the test sets. 
- For both SA and NSA series, fit ARIMA models using all appropriate steps. 
- Measure and discuss how your models perform in-sample and (psuedo-) out-of-sample, comparing candidate models and explaining your choice.
- Fit a polynomial time-trend model to the seasonally-adjusted series and compare its performance to that of your ARIMA model.

### Content
```{r nsa}
# Split series into train and test sets
# Data goes through 2023 W26
nsa_train <- co2_present %>%
  filter(time_index < yearweek("2021 W26"))

nsa_test <- co2_present %>%
  filter(time_index >= yearweek("2021 W26"))

# Fit NSA arima models
# DELETE get order constraint again
nsa_model_aic <- nsa_train %>%
  tsibble::fill_gaps() %>%
  model(ARIMA(average ~ 1 + pdq(1:10,1:2,1:10) + PDQ(0,0,0), ic = "aic", stepwise = FALSE, greedy = FALSE))

nsa_model_aic %>%
  report()

nsa_model_bic <- nsa_train %>%
  tsibble::fill_gaps() %>%
  model(ARIMA(average ~ 1 + pdq(1:10,1:2,1:10) + PDQ(0,0,0), ic = "bic", stepwise = FALSE, greedy = FALSE))

nsa_model_bic %>%
  report()

# Measure performance
```

```{r sa}
# DELETE just difference to get SA?
# sa <- 
```


### Discussion

## How bad could it get?
### Notes
- With the non-seasonally adjusted data series, generate predictions for when atmospheric CO2 is expected to be at 420 ppm and 500 ppm levels for the first and final times (consider prediction intervals as well as point estimates in your answer). 
- Generate a prediction for atmospheric CO2 levels in the year 2122. How confident are you that these will be accurate predictions?

### Content

### Discussion






